---
title: "focal analysis"
format: html
editor: visual
---

Packages

```{r}

library(sf)
library(dplyr)
library(tidyr)
library(ggplot2)
library(scales)
library(patchwork)
library(tmap)
library(gdata)
```

Functions

```{r}

# agr zones
source("../func/calc_agr_zones_sf.R")

# size category
source("../func/site_category.R")

# angle and nn
source("../func/angleCalc.R")
source("../func/nearNeighb.R")

# winkelmass - uniform angle index
source("../func/winkelmass.R")

# differentiation
source("../func/dbhDiff.R")

# dominance
source("../func/dbhDom.R")

# crowding index

# focal overlap index

```

Data

```{r}

st_layers("../data/vect/njs.gpkg")

sites = st_read("../data/vect/njs.gpkg", layer = "sites")

sites = sites %>% 
  mutate(
    xcoord = st_coordinates(sites)[,1],
    ycoord = st_coordinates(sites)[,2]
  )

sites = site_category_numeric(df = sites, col = "size_ha")

```

Plot data by using beginning date for a period

```{r}

ggplot(data = sites, mapping = aes(x = factor(time_start))) +
  geom_bar(fill = "#0b7285") +
  theme_minimal()

```

Select the following periods: 2550BC to 2000BC; 900BC to 600BC, 300BC to 50BC and 300AC to 400AC.

```{r}

sites_2550_BC = sites %>% filter(time_start == -2550)

sites_900_BC =  sites %>% filter(time_start == -900)


```

Map Sites

```{r}
tmap_mode("plot")

map_2550_BC = tm_shape(sites_2550_BC) +
   tm_dots(size = 0.5, alpha = 0.8) +
   tm_layout(main.title = "NJS: 2550BC to 2000BC")

map_900_BC = tm_shape(sites_900_BC) +
   tm_dots(size = 0.5, alpha = 0.8) +
   tm_layout(main.title = "NJS: 900BC to 600BC")


tmap_arrange(map_2550_BC, map_900_BC, nrow = 1)

```
Uniform Angle Index

* 0 - Highly regular
* 0.25 - Regular
* 0.5 - Random
* 0.75 - Clustered
* 1 - Highly Clustered

```{r}

sites_2550_BC$wm = as.factor(winkelmass(sites_2550_BC$xcoord, sites_2550_BC$ycoord, k = 4))

sites_900_BC$wm = as.factor(winkelmass(sites_900_BC$xcoord, sites_900_BC$ycoord, k = 4))

```


```{r}
tmap_mode("plot")

map_2550_BC = tm_shape(sites_2550_BC) +
   tm_dots(size = 0.5, alpha = 0.8, col = "wm", palette = "-GnBu") +
   tm_layout(main.title = "NJS: 2550BC to 2000BC")

map_900_BC = tm_shape(sites_900_BC) +
   tm_dots(size = 0.5, alpha = 0.8, col = "wm", palette = "-GnBu") +
   tm_layout(main.title = "NJS: 900BC to 600BC")


tmap_arrange(map_2550_BC, map_900_BC, nrow = 1)



```
```{r}

w1 = ggplot(data = sites_2550_BC, mapping = aes(x = factor(wm), fill = wm)) +
  geom_bar(aes(y = (..count..)/sum(..count..))) +
  scale_y_continuous(labels=scales::percent) +
  ylab("relative frequencies") +
  scale_fill_brewer(palette = "YlGnBu")+
  labs(title = "NJS: 2550BC to 2000BC") +
  theme_minimal()

w2 = ggplot(data = sites_900_BC, mapping = aes(x = factor(wm), fill = wm)) +
  geom_bar(aes(y = (..count..)/sum(..count..))) +
  scale_y_continuous(labels=scales::percent) +
  ylab("relative frequencies") +
  scale_fill_brewer(palette = "YlGnBu")+
  labs(title = "NJS: 900BC to 600BC") +
  theme_minimal()


w1 + w2

```

Size Dominance

```{r}

# 2550 to 2000

sites_2550_BC_DF = st_drop_geometry(sites_2550_BC)

nb = nearNeighb(sites_2550_BC_DF$xcoord, sites_2550_BC_DF$ycoord, sites_2550_BC_DF$id, k = 4)

sdom = lapply(nb, function(x) {
  nb <- merge(x, sites_2550_BC_DF, by.x = "nb", by.y = "id")
  focal_size <- unique(sites_2550_BC_DF[sites_2550_BC_DF$id == nb$focal,"category"])
  dbhDom(nb$category,focal_size)
  
})

sdom_2550BC = as.data.frame(do.call(rbind, sdom))

sdom_2550BC = sdom_2550BC %>% 
  transmute(
    id = rownames(sdom_2550BC),
    dom = V1
  )

rownames(sdom_2550BC) =  NULL

# 900 to 600

sites_900_BC_DF = st_drop_geometry(sites_900_BC)

nb = nearNeighb(sites_900_BC_DF$xcoord, sites_900_BC_DF$ycoord, sites_900_BC_DF$id, k = 4)

sdom = lapply(nb, function(x) {
  nb <- merge(x, sites_900_BC_DF, by.x = "nb", by.y = "id")
  focal_size <- unique(sites_900_BC_DF[sites_900_BC_DF$id == nb$focal,"category"])
  dbhDom(nb$category,focal_size)
  
})

sdom_900BC = as.data.frame(do.call(rbind, sdom))

sdom_900BC = sdom_900BC %>% 
  transmute(
    id = rownames(sdom_900BC),
    dom = V1
  )

rownames(sdom_900BC) =  NULL

```


Plot size dominance

```{r}

dom1 = ggplot(data = sdom_2550BC, mapping = aes(x = as.factor(dom), fill = as.factor(dom))) +
  geom_bar(aes(y = (..count..)/sum(..count..))) +
  scale_y_continuous(labels=scales::percent) +
  ylab("relative frequencies") +
  scale_fill_brewer(palette = "YlGnBu")+
  labs(title = "NJS: 2550BC to 2000BC") +
  theme_minimal()

dom2 = ggplot(data = sdom_900BC, mapping = aes(x = as.factor(dom), fill = as.factor(dom))) +
  geom_bar(aes(y = (..count..)/sum(..count..))) +
  scale_y_continuous(labels=scales::percent) +
  ylab("relative frequencies") +
  scale_fill_brewer(palette = "YlGnBu")+
  labs(title = "NJS: 900BC to 600BC") +
  theme_minimal()

dom1 + dom2


```


Size Differentiation

```{r}
# 2550 to 2000

sites_2550_BC_DF = st_drop_geometry(sites_2550_BC)

nb = nearNeighb(sites_2550_BC_DF$xcoord, sites_2550_BC_DF$ycoord, sites_2550_BC_DF$id, k = 4)

sdiff = lapply(nb, function(x) {
  nb <- merge(x, sites_2550_BC_DF, by.x = "nb", by.y = "id")
  focal_size <- unique(sites_2550_BC_DF[sites_2550_BC_DF$id == nb$focal,"size_ha"])
  dbhDiff(nb$size_ha,focal_size)
  
})

sdiff_2550BC = as.data.frame(do.call(rbind, sdiff))

sdiff_2550BC = sdiff_2550BC %>% 
  transmute(
    id = rownames(sdiff_2550BC),
    diff = V1
  )

rownames(sdiff_2550BC) =  NULL

# 900 to 600

sites_900_BC_DF = st_drop_geometry(sites_900_BC)

nb = nearNeighb(sites_900_BC_DF$xcoord, sites_900_BC_DF$ycoord, sites_900_BC_DF$id, k = 4)

sdiff = lapply(nb, function(x) {
  nb <- merge(x, sites_900_BC_DF, by.x = "nb", by.y = "id")
  focal_size <- unique(sites_900_BC_DF[sites_900_BC_DF$id == nb$focal,"size_ha"])
  dbhDiff(nb$size_ha,focal_size)
  
})

sdiff_900BC = as.data.frame(do.call(rbind, sdiff))

sdiff_900BC = sdiff_900BC %>% 
  transmute(
    id = rownames(sdiff_900BC),
    diff = V1
  )

rownames(sdiff_900BC) =  NULL
```

Plot Size Differentation

```{r}

size_diff = combine(sdiff_2550BC,sdiff_900BC)

ggplot() +
  geom_density(data = size_diff, aes(x = diff, color = source), alpha = 0.3) +
  scale_color_manual(values=c("#9ecae1","#fdae6b")) +
  geom_vline(aes(xintercept = mean(sdiff_2550BC$diff)), color = "#9ecae1", linetype = "dotted") +
  geom_vline(aes(xintercept= mean(sdiff_900BC$diff)), color = "#fdae6b", linetype = "dotted") +
  labs(title = "Size Differentiation",
       subtitle = "0 implies neighbouring sites have equal size") +
  theme_minimal()

```

Crowding Index A - based on Agricultural Zones

```{r}

# 2550BC calculate agr zones
sites_2550_BC_AGR = calc_agr_zones_sf(sites_2550_BC, size = "size_ha")

# calculate number of overlapping agr zones
sites_2550_BC_AGR$n.overlap = lengths(st_overlaps(sites_2550_BC_AGR))

# calculate index 

sites_2550_BC_AGR$focal.over = sites_2550_BC_AGR$n.overlap / 4

# 900BC calculate agr zones
sites_900_BC_AGR = calc_agr_zones_sf(sites_900_BC, size = "size_ha")

# calculate number of overlapping agr zones
sites_900_BC_AGR$n.overlap = lengths(st_overlaps(sites_900_BC_AGR))

sites_900_BC_AGR$n.overlap = replace(sites_900_BC_AGR$n.overlap, sites_900_BC_AGR$n.overlap>4, 4)

# calculate index 
sites_900_BC_AGR$focal.over = sites_900_BC_AGR$n.overlap / 4





```
Plot

```{r}
f1 = ggplot(data = sites_2550_BC_AGR, mapping = aes(x = as.factor(focal.over), fill = as.factor(focal.over))) +
  geom_bar(aes(y = (..count..)/sum(..count..))) +
  scale_y_continuous(labels=scales::percent) +
  ylab("relative frequencies") +
  scale_fill_brewer(palette = "YlGnBu")+
  labs(title = "NJS: 2550BC to 2000BC") +
  theme_minimal()

f2 = ggplot(data = sites_900_BC_AGR, mapping = aes(x = as.factor(focal.over), fill = as.factor(focal.over))) +
  geom_bar(aes(y = (..count..)/sum(..count..))) +
  scale_y_continuous(labels=scales::percent) +
  ylab("relative frequencies") +
  scale_fill_brewer(palette = "YlGnBu")+
  labs(title = "NJS: 900BC to 600BC") +
  theme_minimal()

f1 + f2
```
Crowding Index B - based on proportion of Agricultural Zones Area overlapping

```{r}

# 2550BC


s2550BC = sites_2550_BC_AGR %>% 
  st_intersection() %>% 
  filter(n.overlaps == 1) %>% 
  mutate(
    area_free = unclass(st_area(.)),
    area_over = agr_area - area_free,
    index = area_over / agr_area
  )


# 900BC
s900BC = sites_900_BC_AGR %>% 
  st_intersection() %>% 
  filter(n.overlaps == 1) %>% 
  mutate(
    area_free = unclass(st_area(.)),
    area_over = agr_area - area_free,
    index = area_over / agr_area
  )

  


```

Map
```{r}

tmap_mode("view")
tmap_options(check.and.fix = TRUE)
st_is_valid(s2550BC)
tm_shape(s2550BC) +
  tm_polygons(col = "index") +
  tm_shape(s900BC) +
  tm_polygons(col = "index") 
  
  


```

Plot

```{r}
crowd = combine(s2550BC,s900BC)

ggplot() +
  geom_density(data = crowd, aes(x = index, color = source), alpha = 0.3) +
  scale_color_manual(values=c("#9ecae1","#fdae6b")) +
  geom_vline(aes(xintercept = mean(s2550BC$index)), color = "#9ecae1", linetype = "dotted") +
  geom_vline(aes(xintercept= mean(s900BC$index)), color = "#fdae6b", linetype = "dotted") +
  labs(title = "Crowding Index",
       subtitle = "0 implies no overlap - 1 implies total overlap") +
  theme_minimal()

# Look at proportions
```


